---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README",
  out.width = "100%"
)

options(width = 80)
```

# parsermd

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of parsermd is to extract the content of an Rmarkdown file to allow for programmatic interactions with the code chunks and markdown content.

## Installation

You can install the released version of parsermd from [GitHub](https://github.com/rundel/parsermd) with:

```r
remotes::install_github("rundel/parsermd")
```

```{r}
library(parsermd)
```

## Example

This is a basic example which shows you the basic abstract syntax tree (AST) that results from parsing a simple Rmd file,

```{r example}
rmd = parsermd::parse_rmd(system.file("minimal.Rmd", package = "parsermd"))
```

The RMarkdown document is parsed and stored in a flat, ordered list object containing tagged elements. By default the package will present a heirachical view of the document where chunks and markdown text are nested within headings, this is shown by the default print methods.

```{r tree}
print(rmd)
```

If you would prefer to see the underlying flat structure, this can be seen by setting `use_headings = FALSE` with `print`.

```{r no_headings}
print(rmd, use_headings = FALSE)
```

Additionally, to ease the manipulation of the AST the package supports the transformation of the object into a tidy tibble with `as_tibble` or `as.data.frame` (both return a tibble).

```{r tibble}
as_tibble(rmd)
```

and as expected this data frame representation can also be converted back into an AST.

```{r as_ast}
as_ast( as_tibble(rmd) )
```

Finally, we can also convert the AST back into an RMarkdown document via `as_document`

```{r as_doc}
cat(
  as_document(rmd),
  sep = "\n"
)
```


